The final layer of a classifier is stochastic: we form a softmax distribution over the classes and sample from it. We ask what happens when we do the same thing one layer earlier --- treat an intermediate layer's activations as a distribution and draw a sample, instead of passing the deterministic activations forward. Hidden-layer stochasticity is the shared mechanism behind variational autoencoders, Gumbel-Softmax latents, stochastic neurons, dropout, and the information bottleneck, but each of these studies it for a purpose or at a fixed place; no prior work isolates intermediate-layer activation sampling as the variable of study, and the analogy to sampling the output softmax is rarely made explicit. We run that controlled study with a single drop-in sampling layer, holding architecture, data, and optimizer fixed while varying only the activation operator across seven choices: deterministic, dropout, additive Gaussian noise, continuous Gaussian reparameterization, Bernoulli straight-through, and categorical Gumbel-Softmax in soft and hard forms. Across MNIST (MLP) and CIFAR-10 (CNN), 3 seeds and 117 trained models, we measure accuracy, NLL, calibration, generalization gap, gradient variance, MC-averaging, and sweeps over temperature, layer position, and input noise. We find that intermediate-layer sampling is a distribution-type-dependent regularizer. Continuous Gaussian sampling is essentially free (97.7\% vs 97.8\% on MNIST) and on CIFAR-10 beats the deterministic baseline on accuracy (75.5\% vs 75.0\%), NLL (0.78 vs 0.94), and calibration (ECE 0.074 vs 0.112). The literal softmax analogue is the opposite: categorical sampling costs 8--10 points on MNIST and collapses to chance (10\%) on CIFAR-10, and its gradient variance is the highest of all operators, exactly tracking the accuracy ranking. MC-averaging recovers most of the single-sample cost, mirroring how averaging softmax samples approximates the expectation. Sampling a hidden layer like the output softmax is therefore possible and even beneficial in its gentle continuous form, but the literal categorical form is a severe information bottleneck whose cost scales with task difficulty.
